---
title: "Inference with many instruments"
author: "Michal Kolesár"
date: "`r Sys.Date()`"
bibliography: many-iv-library.bib
output:
  pdf_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Inference with many instruments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE, cache=FALSE}
library("knitr")
knitr::opts_knit$set(self.contained = FALSE)
knitr::opts_chunk$set(tidy = TRUE, collapse=TRUE, comment = "#>",
                      tidy.opts=list(blank=FALSE, width.cutoff=55))
```

# Summary

The package `RDHonest` implements estimators and confidence intervals in a
linear instrumental variables model considered in @kolesar17md and @kcfgi14. In
this vignette, we demonstrate the implementation of these estimators and
confidence intervals using a subset of the dataset used in @ak91, which is
included in the package as a data frame `ak80`. This data frame corresponds to a
sample of males born in the US in 1930--39 from 5% sample of the 1980 Census.
See `help("ManyIV::ak80")` for details.

# Estimation and Inference

The package implements the following estimators via the command `IVreg`

1. Two-stage least-squares (TSLS) estimator
2. Limited information maximum likelihood (LIML) estimator  due to @ar49
3. A modification of the bias-corrected two-stage least squares (MBTSLS)
   estimator (@kcfgi14) that slightly modifies the original @nagar59 estimator
   so that it's consistent under many exogenous regressors as well as many
   instruments.
4. Efficient minimum distance (EMD) estimator [@kolesar17md] that is more
   efficient than LIML under many instrument asymptotics unless the reduced-form
   errors are Gaussian.

`IVreg` computes the following types of standard errors are computed:

1. Conventional homoscedastic standard errors, as computed by Stata's
   `ivregress` and `ivreg2`. These standard errors are not robust to many
   instruments (option `inference=standard`)
2. Conventional heteroscedastic standard errors, as computed by Stata's
   `ivregress` and `ivreg2`. These standard errors are not robust to many
   instruments. (option `inference=standard`)
3. Standard errors based on the information matrix of the limited information
   likelihood of @ar49 (for LIML only). These are not robust to many instruments
   (option `inference=lil`)
3. Standard errors based on the Hessian of the random-effects likelihood
   of @ci04. These standard errors are for LIML only (since the random-effects ML
   estimator coincides to LIML), and are robust to many instruments provided the
   reduced-form errors are Gaussian (option `inference=re`).
4. Standard errors based on the Hessian of the invariant
   likelihood [see @kolesar17md]. These standard errors are for LIML only (since
   the invariant ML estimator coincides to LIML), and are robust to many
   instruments provided the reduced-form errors are Gaussian. This involves some
   numerical optimization. (option `inference=il`)
5. Many-instrument robust standard errors based on the minimum distance
   objective function [see @kolesar17md] (option `inference=md`). Since the TSLS
   estimator is not consistent under many-instrument asymptotics, its standard
   errors are omitted. Unlike the `re` and `il` standard errors, the standard
   errors for MBTSLS, LIML and EMD do not require the reduced-form errors to be
   Gaussian. In addition, the command computes standard errors for MBTSLS based
   on the unrestricted minimum distance objective function (`umd`), which allows
   for treatment effect heterogeneity (provided the reduced-form errors remain
   homoscedastic), and for failures of the exclusion restriction as considered
   in @kcfgi14.

Several of these options may be specified at once:

```{r}
library("ManyIV")
## Specification as in Table V, columns (1) and (2) in Angrist and Krueger
IVreg(lwage~education+as.factor(yob)|as.factor(qob)*as.factor(yob),
            data=ak80, inference=c("standard", "re", "il", "lil"))
```

With large data, the `md` standard errors may take a while to run for large data
sets, as they require estimation of third and fourth moments of the reduced-form
errors.

[[TODO EMD]]

## Specification testing

The command

# Implementation details

\newcommand{\by}{\mathbf{Y}} %
\newcommand{\bx}{\mathbf{X}} %
\newcommand{\bg}{\mathbf{G}} %
\newcommand{\bz}{{Z}} %
\newcommand{\bp}{\mathbf{P}} %
\newcommand{\br}{\mathbf{R}} %
\newcommand{\bw}{{W}} %
\newcommand{\eye}{{I}} %
\newcommand{\Ni}{K} %
\newcommand{\Ne}{L} %
\newcommand{\Hm}[1]{{H}_{#1}} %hat matrix
\newcommand{\Dm}[1]{\mathbf{D}_{#1}} %diagonal matrix
\newcommand{\tsls}{\text{\textsc{tsls}}}%
\newcommand{\ols}{\text{\textsc{ols}}}%
\newcommand{\umd}{\text{\textsc{umd}}}%
\newcommand{\re}{\text{\textsc{re}}}%
\newcommand{\mbtsls}{\text{\textsc{mbtsls}}}%
\newcommand{\liml}{\text{\textsc{liml}}}%
\newcommand{\h}[2]{\hat{#1}_{\text{\textsc{#2}}}} %

Let $$y_{i}=x_{i}\beta+W_{i}'\delta+\epsilon_{i},$$ where $y_{i}\in\mathbb{R}$
is the outcome variable, $x_{i}\in\mathbb{R}$ is a single endogenous regressor,
$W_{i}\in\mathbb{R}^{\ell}$ is a vector of exogenous regressors (covariates),
and $\epsilon_{i}$ is a structural error. The parameter of interest is $\beta$.
In addition, $Z_{i}\in\mathbb{R}^{k}$ is a vector of instruments.

We observe an i.i.d.~sample $\{y_{i},x_{i},W_{i},Z_{i}\}_{i=1}^{n}$. The
arguments of `ivreg` are the matrices $Y$, $Z$, and $W$, with
rows $(y_{i},x_{i})$, $Z_{i}'$ and $W_{i}'$. The matrix $Z$ doesn't necessarily
have orthonormal columns.

For any full-rank $n\times m$ matrix ${A}$, let $\Hm{{A}}={A}({A}'{A})^{-1}{A}'$
denote the associated $n\times n$ projection matrix (also known as the hat
matrix). Let $\eye_{m}$ denote the $m\times m$ identity matrix. Let
${A}_{\perp}=(\eye_{n}-\Hm{W}){A}$ denote the residual from the sample projection of
${A}$ onto $\bw$.

Define matrices $S$ and $T$ as in MD paper:
\begin{align*}
  T&=Y_{\perp}'\Hm{\bz_{\perp}}Y_{\perp}/n,&
  S&=Y'(\eye_{n}-\Hm{\bz,\bw})Y/(n-\Ni-\Ne).
\end{align*}
Also define $m_{\min}$ and $m_{\max}$ to be the minimum and maximum eigenvalues of the matrix $S^{-1}T$. A $k$-class estimator estimator with parameter $\kappa$ is then given by
\begin{equation*}
  \hat{\beta}(\kappa)=\frac{T_{1,2}-m(\kappa) S_{2,2}}{T_{2,2}-m(\kappa)S_{22}},
\end{equation*}
where $m(\kappa)=(\kappa-1)(1-\Ni/n-\Ne/n)$, so that
\begin{align*}
  m_{\ols}&=-(1-\Ni/n-\Ne/n)&
  m_{\tsls}&=0,& m_{\mbtsls}&=k/n,& m_{\liml}&=m_{\min}.
\end{align*}

## Stata standard errors
Stata 13's =ivregress= and =ivreg2= use standard errors for
$k$-class estimators given by
\begin{equation*}
  \widehat{var}_{\text{Stata}}(\hat{\beta}(\kappa))=
  \frac{\hat{\sigma}(\kappa)^{2}}{(1-\kappa)x_{\perp}'x_{\perp}+\kappa x'\Hm{Z_{\perp}}x}
  =  \frac{1}{n}\frac{\hat{\sigma}(\kappa)^{2}}{
    T_{2,2}-m(\kappa) S_{2,2}},
\end{equation*}
where
$\hat{\sigma}(\kappa)^{2}=\hat{\epsilon}(\kappa)'\hat{\epsilon}(\kappa)/n$, with
$\hat{\epsilon}(\kappa)=y_{\perp}-x_{\perp}\hat{\beta}(\kappa)=y-x\hat{\beta}(\kappa)-W'\hat{\delta}(\kappa)$,
and $\hat{\delta}(\kappa)=(\bw'\bw)^{-1}\bw'(y-x\hat{\beta}(\kappa))$. This
includes \liml, for which $\kappa$ is random (Stata disregards that). For \ols,
we use the Stata 13 variance estimator
$\hat{\sigma}=\hat{\epsilon}_{\ols}'\hat{\epsilon}_{\ols}/(n-\Ne-1)$.

To define the robust standard error estimators, let
$\hat{R}_{i}=Z_{\perp,i}(Z_{\perp}'Z_{\perp})^{-1}Z_{\perp}x$. Then, for a
$k$-class estimator (including \liml),
\begin{equation*}
  \widehat{var}_{\text{StataR}}(\hat{\beta}(\kappa))
  =\frac{\sum_{i=1}^{n}\hat{\epsilon}_{i}(\kappa)\hat{R}_{i}^{2}}{n^{2}(T_{2,2}-m(\kappa)S_{2,2})^{2}}.
\end{equation*}
Note that $\widehat{var}_{\text{StataR}}(\hat{\beta}(\kappa))$ and
$\widehat{var}_{\text{Stata}}(\hat{\beta}(\kappa))$ don't necessarily converge
to the same quantity even under homoskedasticity. For \ols, we use
$(n/(n-\Ne-1))^{1/2}x_{\perp}$ in place of $\hat{R}_{i}$.

One could alternatively use $T_{2,2}$ in the denominator, or estimate
$var(\epsilon_{i})$ using $\hat{\sigma}(\beta)=(1,-\beta)S(1,-\beta)'$. Such
variance estimators were used in invalid IV paper. The alternative denominator
makes a huge difference. For the AK example in invalid IV paper, we have
$T_{2,2}-(k/n)S_{2,2}=0.0078$, but $T_{2,2}=0.03905$. How we estimate
$\sigma^{2}$ matters less, $\hat{\sigma}^{2}(\kappa)=0.4021$, and
$(1,-\hat{\beta})S(1,-\hat{\beta})=0.4034$. In this dataset, $S$ is two orders
of magnitude bigger than $T$, and with $k/n=0.003$, $T$ and $k/n S$ are same
order of magnitude, so whether we subtract it or not makes a huge difference.

## Likelihood-based standard errors
For \liml, we also compute standard errors based on the information matrix of
the limited information likelihood (see Equation (SA-5) in MD paper),
\begin{equation*}
  \widehat{var}_{\liml}(\hat{\beta}_{\liml})=\frac{\hat{a}_{\liml}'
    \hat{\Omega}_{\liml}^{-1}\hat{a}_{\liml}\cdot \hat{b}_{\liml}'
    \hat{\Omega}_{\liml}\hat{b}_{\liml} }{\hat{\lambda}_{\liml}},
\end{equation*}
where we use the \liml\ estimators
\begin{align*}
  \h{\Omega}{liml}&= \frac{n-\Ne-\Ni}{n-\Ne}S+\frac{n}{n-\Ne}\left(T-
    \frac{m_{\max}\h{a}{liml}\h{a}{liml}'}{\h{a}{liml}'S^{-1}\h{a}{liml}}\right),\\
  \h{\lambda}{liml}&=\frac{n-\Ne}{n-k-\Ne}m_{\max}.
\end{align*}
We also compute standard errors based on the Hessian of the \re\ likelihood,
    \begin{equation*}
      \widehat{var}_{\re}(\h{\beta}{liml})=-n \h{\mathcal{H}}{re}^{11}=
      -\frac{\hat{b}'\h{\Omega}{re}\hat{b}
        (\h{\lambda}{re}+k/n)}{\h{\lambda}{re}}\left(
        \hat{Q}_{\mathcal{S}}\h{\Omega}{re,22}-T_{22}+\frac{\hat c}{1-\hat c}\frac{\hat{Q}_{\mathcal{S}}}{\hat{a}'
          \h{\Omega}{re}^{-1}\hat{a}} \right)^{-1},
    \end{equation*}
where
    \begin{align}
      \h{\lambda}{re}&=m_{\max}-\Ni /n,\\
      \h{\Omega}{re}&=\frac{n-\Ni-\Ne}{n-\Ne}S+\frac{n}{n-\Ne}\left(T-
        \frac{\h{\lambda}{re}}{\hat{a}'S^{-1}\hat{a} } \hat{a}\hat{a}'\right),&
      \hat{a}&=
      \begin{pmatrix}
        \hat{\beta}_{\liml}\\1
      \end{pmatrix},
      \\
      \hat{Q}_{\mathcal{S}} &=\frac{\hat{b}'S \hat{b}}{\hat{b}'\h{\Omega}{re} \hat{b}},& \hat{b}
      &=
  \begin{pmatrix}
    1\\-\hat{\beta}_{\liml}
  \end{pmatrix},\\
\hat c&=\frac{\h{\lambda}{re} \hat{Q}_{\mathcal{S}}}{(K/n+\h{\lambda}{re})(1-\Ne/n)}.
    \end{align}
    The estimator for the standard error of \mbtsls\ is based on the
    Hessian of the uncorrelated random effects likelihood with the estimator of
    the variance of the direct effects set to zero,
    \begin{equation*}
      \h{\mathcal{H}}{ure}=
      \frac{1}{\hat{\Lambda}_{22}^{2}}\left(\hat{\Lambda}_{22}\hat{\Sigma}_{11}+
        \frac{(1-\Ne/n )k/n}{(1-\Ni/n-\Ne/n)}
        (\hat{\Sigma}_{11}\hat{\Sigma}_{22}+\hat{\Sigma}_{12}^{2})    \right),
    \end{equation*}
where
\begin{align*}
  \hat{\Lambda}_{22}&= \begin{cases} S_{22}-\frac{K }{n}\h{\Omega}{22,ure} &
    \text{if $m_{\min}\geq K/n$,} \\
    \frac{\h{\lambda}{re}}{\h{a}{re}'\h{\Omega}{re}^{-1}\h{a}{re}} &
    \text{otherwise.}
      \end{cases},\\
      \h{\Omega}{ure}&=\begin{cases}
        S & \text{if $m_{\min}\geq K/n$,} \\
        \h{\Omega}{re} & \text{otherwise.}
      \end{cases},\\
      \h{\Sigma}{ure} &= \h{\Gamma}{ure}'\h{\Omega}{ure}\h{\Gamma}{ure}, &
      \h{\Gamma}{ure} &=\begin{pmatrix} 1& 0\\
        -\hat{\beta}_{\mathtt{mbtsls}} &1
  \end{pmatrix}.
\end{align*}

## Many instruments plug-in standard errors
Under Normality, the standard errors for \liml\ and \mbtsls\ under many
instruments are given by
\begin{equation*}
avar=  \frac{1}{\Lambda^{2}}\left(\Lambda\Sigma_{11}+\frac{(1-\Ne/n)k/n}{
      1-\Ni/n-\Ne/n}(\det(\Sigma)+
    2d\Sigma_{12})\right),
\end{equation*}
where $d=0$ for \liml, and $d=1$ for \mbtsls. We estimate them using the plug-in
estimates based on \re\ estimates or \umd\ estimates of $\Lambda$, $\beta$, and
$\Omega$.

[[TODO: Non-normal errors for liml and mbtsls based on MD objective]]

## Many invalid instruments
The estimator for the standard error of \texttt{mbtsls} is based on the Hessian
of the uncorrelated random effects likelihood,
    \begin{equation*}
      \h{\mathcal{H}}{ure}=
      \frac{1}{\hat{\Lambda}_{22}^{2}}\left(\hat{\Lambda}_{22}\hat{\Sigma}_{11}+
        \frac{(1-\ell/n )k/n}{(1-k/n-\ell/n)}
        (\hat{\Sigma}_{11}\hat{\Sigma}_{22}+\hat{\Sigma}_{12}^{2})
        +\hat{\Lambda}_{11}\hat{\Sigma}_{22}+
        \hat{\Lambda}_{11}\Lambda_{22}\cdot n/K
      \right),
    \end{equation*}
    where $\hat{\Lambda}_{22}$ and $\hat{\Sigma}_{22}$ are computed as before,
    and $\hat{\Lambda}_{11}=\max\left\{\hat{b}_{\mathtt{mbtsls}}'(S-\frac{K
        }{n}S_{\perp}) \hat{b}_{\mathtt{mbtsls}} ,0\right\}.$

## Other outputs
In addition, we compute the first-stage $F$-statistic,
\begin{equation*}
F= \frac{n}{k}\frac{T_{22}}{S_{22}}.
\end{equation*}
We also compute the Sargan test statistic and p-value,
$nm_{\min}/(1-p/n-\ell/n+m_{\min}))$, using $\chi^{2}_{k-1}$. The Sargan test
statistic is based on \liml, unlike in Stata 13's =estat overid=, where
it depends on what estimator was used to compute $\beta$.

We also compute the test of overidentifying restrictions derived in MD paper.
The $p$-value under Normality is given by
\begin{equation*}
  1-\Phi(\sqrt{(n-k-\ell)/(n-\ell)}\Phi^{-1}(F_{k-1}(nm_{\min}))),
\end{equation*}
where $\Phi$ is the cdf of a standard Normal distribution, and $F_{k-1}$ is the
cdf of a $\chi^{2}$ distribution with $k-1$ degrees of freedom.




# References
