---
title: "Inference with Many Instruments"
author: "Michal Kolesár"
date: "`r Sys.Date()`"
bibliography: many-iv-library.bib
output:
  pdf_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Inference with many instruments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE, cache=FALSE}
library("knitr")
knitr::opts_knit$set(self.contained = FALSE)
knitr::opts_chunk$set(tidy = TRUE, collapse=TRUE, comment = "#>",
                      tidy.opts=list(blank=FALSE, width.cutoff=55))
```

# Summary

The package `ManyIV` implements estimators and confidence intervals in a linear
instrumental variables model considered in @kolesar18md and @kcfgi14. In this
vignette, we demonstrate the implementation of these estimators and confidence
intervals using a subset of the dataset used in @ak91, which is included in the
package as a data frame `ak80`. This data frame corresponds to a sample of males
born in the US in 1930--39 from 5% sample of the 1980 Census. See
`help("ManyIV::ak80")` for details.

# Estimation and Inference

The package implements the following estimators via the command `IVreg`

1. Two-stage least-squares (TSLS) estimator
2. Limited information maximum likelihood (LIML) estimator  due to @ar49.
3. A modification of the bias-corrected two-stage least squares (MBTSLS)
   estimator (@kcfgi14) that slightly modifies the original @nagar59 estimator
   so that it's consistent under many exogenous regressors as well as many
   instruments, provided the reduced-form errors are homoskedastic.
4. Efficient minimum distance (EMD) estimator (@kolesar18md) that is more
   efficient than LIML under many instrument asymptotics unless the reduced-form
   errors are Gaussian.

`IVreg` computes the following types of standard errors:

1. Conventional homoskedastic standard errors, as computed by Stata's
   `ivregress` and `ivreg2`. These standard errors are not robust to many
   instruments (option `inference=standard`)
2. Conventional heteroskedastic standard errors, as computed by Stata's
   `ivregress` and `ivreg2`. These standard errors are not robust to many
   instruments. (option `inference=standard`)
3. Standard errors that are valid under heterogeneous treatment effects as well
   as heteroskedasticity (labeled `HTE robust`). These standard errors are not
   robust to many instruments (option `inference=standard`). They are only
   computed for TSLS and MBTSLS, since LIML is not robust to heterogeneous
   treatment effects (see @kolesar13late).
4. Standard errors based on the information matrix of the limited information
   likelihood of @ar49 (for LIML only). These are not robust to many instruments
   or heteroskedasticity (option `inference=lil`)
5. Standard errors based on the Hessian of the random-effects likelihood of @ci04.
   These standard errors are for LIML only (since the random-effects ML
   estimator coincides to LIML), and are robust to many instruments provided the
   reduced-form errors are Gaussian and homoskedastic (option `inference=re`).
6. Standard errors based on the Hessian of the invariant likelihood (see
   @kolesar18md). These standard errors are for LIML only (since the invariant
   ML estimator coincides to LIML), and are robust to many instruments provided
   the reduced-form errors are Gaussian and homoskedastic. This involves some
   numerical optimization. (option `inference=il`)
7. Many-instrument robust standard errors based on the minimum distance
   objective function (see @kolesar18md) (option `inference=md`). Since the TSLS
   estimator is not consistent under many-instrument asymptotics, its standard
   errors are omitted. Unlike the `re` and `il` standard errors, the standard
   errors for MBTSLS, LIML and EMD do not require the reduced-form errors to be
   Gaussian, although the homoskedasticity assumption is still needed. In
   addition, the command computes standard errors for MBTSLS based on the
   unrestricted minimum distance objective function (`umd`), which allows for
   treatment effect heterogeneity (provided the reduced-form errors remain
   homoskedastic), and for failures of the exclusion restriction as considered
   in @kcfgi14.

Several of these options may be specified at once:
```{r}
library("ManyIV")
## Specification as in Table V, columns (1) and (2) in Angrist and Krueger
IVreg(lwage~education+as.factor(yob)|as.factor(qob)*as.factor(yob),
            data=ak80, inference=c("standard", "re", "il", "lil"))
```

With large data, the `md` standard errors may take a while to run, as they
require estimation of third and fourth moments of the reduced-form errors. In
particular, letting $M$ denote the annihilator matrix associated with the matrix
$(W,Z)$ of exogenous regressors and instruments, the formulas for these moments
require the computation of $\tilde{m}_{3}=\sum_{i,j}M_{i,j}^3$, and
$\tilde{m}_{4}=\sum_{i,j}M_{i,j}^4$. If option `approx=TRUE` is selected (which
is the default), to speed up the calculations, the function `ivreg` uses the
approximation $\tilde{m}_{3}\approx n-3(k+l)$ and $\tilde{m}_{4}\approx
n-4(k+l)$, where $n$ is the sample size, $k$ is the number of instruments, and
$\ell$ is the number of exogenous regressors. This approximation is accurate up
to terms of order $O((k+l)/n)^2)$, and should have a negligible effect on the
estimates unless the ratio $(k+l)/n$ is quite large. With this approximation,
the calculations are quite fast even for large sample sizes:

```{r}
r1 <- IVreg(lwage~education+as.factor(yob)|as.factor(qob)*as.factor(yob),
            data=ak80, inference="md", approx=TRUE)
print(r1, digits=4)
```

We can see that the LIML and EMD estimates are identical up to 4 significant
digits.

## Specification testing

The package also implements two tests for overidentifying restrictions. The
first test is the classic @sargan58 test. The second test is a modification of
the @cd93 test developed in @kolesar18md to make the test robust to many
instruments and many exogenous regressors (provided the reduced-form errors are
homoskedastic). The command `IVoverid` takes the results of the IV regression as
an argument.

```{r}
IVoverid(r1)
```

# Implementation details

\newcommand{\Hm}[1]{{H}_{#1}}


Let $$y_{i}=x_{i}\beta+w_{i}'\delta+\epsilon_{i},$$ where $y_{i}\in\mathbb{R}$
is the outcome variable, $x_{i}\in\mathbb{R}$ is a single endogenous regressor,
$w_{i}\in\mathbb{R}^{\ell}$ is a vector of exogenous regressors (covariates),
and $\epsilon_{i}$ is a structural error. The parameter of interest is $\beta$.
In addition, $z_{i}\in\mathbb{R}^{k}$ is a vector of instruments.

We observe an i.i.d.~sample $\{y_{i},x_{i},w_{i},z_{i}\}_{i=1}^{n}$. Let $Y$,
$Z$, and $W$, denote matrices with rows $(y_{i},x_{i})$, $z_{i}'$ and $w_{i}'$.
For any full-rank $n\times m$ matrix ${A}$, let $\Hm{{A}}={A}({A}'{A})^{-1}{A}'$
denote the associated $n\times n$ projection matrix (also known as the hat
matrix). Let $I_{m}$ denote the $m\times m$ identity matrix, and let
${Z}_{\perp}=(I_{n}-\Hm{W}){Z}$ denote the residual from the sample
projection of ${Z}$ onto $W$.

Define matrices $S$ and $T$ as in @kolesar18md: \begin{align*}
T&=Y'\Hm{Z_{\perp}}Y/n,& S&=Y'(I_{n}-\Hm{Z,W})Y/(n-k-\ell). \end{align*}
Also define $m_{\min}$ and $m_{\max}$ to be the minimum and maximum eigenvalues
of the matrix $S^{-1}T$. The estimators TSLS, OLS, MBTSLS, and LIML are all
$k$-class estimators. A $k$-class estimator estimator with parameter $\kappa$ is
then given by
\begin{equation*}
\hat{\beta}(\kappa)=\frac{T_{12}-m(\kappa)
S_{12}}{T_{22}-m(\kappa)S_{22}},
\end{equation*}
where $m(\kappa)=(\kappa-1)(1-k/n-\ell/n)$. For the estimators above,
\begin{align*}
m_{OLS}&=-(1-k/n-\ell/n)& m_{TSLS}&=0,& m_{MBTSLS}&=k/n,&
m_{LIML}&=m_{\min}.
\end{align*}
The EMD estimator is not a $k$-class estimator.

The `li`, `lil`, `re`, and `md` standard errors are based on the formulas
described in @kolesar18md. In the remainder of this vignette, we briefly
describe the formulas for conventional standard errors.

## Other standard errors
Stata 13's `ivregress` and `ivreg2` use standard errors for
$k$-class estimators given by
\begin{equation*}
  \widehat{var}_{\text{Stata}}(\hat{\beta}(\kappa))
  =  \frac{1}{n}\frac{\hat{\sigma}(\kappa)^{2}}{
    T_{22}-m(\kappa) S_{22}},
\end{equation*}
where
$\hat{\sigma}(\kappa)^{2}=\hat{\epsilon}(\kappa)'\hat{\epsilon}(\kappa)/n$, with
$\hat{\epsilon}(\kappa)=y-x\hat{\beta}(\kappa)-W'\hat{\delta}(\kappa)$,
and $\hat{\delta}(\kappa)=(W' W)^{-1} W'(y-x\hat{\beta}(\kappa))$. This
includes LIML, for which $\kappa$ is random (Stata disregards that). For OLS,
we use the Stata 13 variance estimator
$\hat{\sigma}=\hat{\epsilon}_{OLS}'\hat{\epsilon}_{OLS}/(n-\ell-1)$.

To define the robust standard error estimators, let
$\hat{R}_{i}=Z_{\perp, i}(Z_{\perp}'Z_{\perp})^{-1}Z_{\perp}x$. Then, for a
$k$-class estimator (including LIML),
\begin{equation*}
  \widehat{var}_{\text{Stata, robust}}(\hat{\beta}(\kappa))
  =\frac{\sum_{i=1}^{n}\hat{\epsilon}_{i}(\kappa)^2\hat{R}_{i}^{2}}{n^{2}(T_{22}-m(\kappa)S_{22})^{2}}.
\end{equation*}
Note that $\widehat{var}_{\text{Stata, robust}}(\hat{\beta}(\kappa))$ and
$\widehat{var}_{\text{Stata}}(\hat{\beta}(\kappa))$ don't necessarily converge
to the same quantity even under homoskedasticity. For OLS, we use
$(n/(n-\ell-1))^{1/2}x_{\perp}$ in place of $\hat{R}_{i}$.

One could alternatively use $T_{22}$ in the denominator, or estimate
$var(\epsilon_{i})$ using $\hat{\sigma}(\beta)=(1,-\beta)S(1,-\beta)'$. Such
variance estimators were used in @kcfgi14. The alternative denominator makes a
big difference, but how we estimate $\sigma^{2}$ matters less.

## Other outputs
The first-stage $F$-statistic reported by `IVreg` is given by
\begin{equation*}
F= \frac{n}{k}\frac{T_{22}}{S_{22}}.
\end{equation*}

The Sargan test statistic is given by $n m_{\min}/(1-p/n-\ell/n+m_{\min}))$, and
its $p$-value is based on a $\chi^{2}_{k-1}$ approximation. The Sargan test
statistic is based on LIML, unlike in Stata 13's `estat overid`, where it
depends on what estimator was used to compute $\beta$. The adjusted Cragg-Donald
test is described in @kolesar18md [Section 6].

# References
